Parallel-in-time Methods for Particle Simulations

### 2025-02-06 paper/main.tex
Fixed links
Paper improvements

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{comment}
\usepackage[thinc]{esdiff}
\usepackage{xurl}

@@

\author{\IEEEauthorblockN{Lukáš Veškrna}
\IEEEauthorblockA{\textit{Technical University Munich}\\ \textit{\& Charles University} \\
Munich, Germany \\
lukas.veskrna@gmail.com}
}

@@

\begin{abstract}
The purpose of this paper is to examine the behavior of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator in the field of solar system dynamics utilizing parallel-in-time integration. We present the results of experiments measuring the run times of the simulation, which we then compare with expectations derived from theory.
\end{abstract}

@@

If we want to learn more about the behavior of mostly stable planetary orbits, we have to simulate a long time. Thus, parallel-in-time algorithms can be employed. \cite{parallelsolar}

@@

We typically solve this problem using a numerical method, that does a step in time of size $\Delta t$ and estimates $u(t_0+\Delta t)$. We then use this result to advance the computation further in time by taking more such steps.

@@

Supposing we integrate from time $t_0$ to $T$, we split this interval into $N$ segments of size $\Delta t = \frac{T-t_0}{N}$. We denote $u_0, u_1\dots u_N$ the states of the system in times $t_0, t_1\dots t_N$ with $t_n = t_0 + n\Delta t$ (so $t_0 = t_0$ and $t_N = T$).\footnote{The solvers can also use their own smaller subdivisions (e.g. different step sizes) but the parareal algorithm does not interact with those subdivisions.}

@@

This is repeated until satisfactory convergence. It has been proven for example by \cite{parareal2} that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta t, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.

This method has been formulated using different generalized frameworks in \cite{parareal2}. They also extend the time interval paritioning to arbitrary partions, not only equally sized ones as in \cite{parareal}.

@@

A derivation of the theoretical speedup is presented in Pentland et al. \cite{speedup} (Section 2.3), which we use here. It calculates the speed-up when running even the coarse solver partially in parallel, we simplify it with running $\mathcal{C}$ serially (our implementation doesn't include this optimization).

@@

It is also possible to use a similar algorithm to the parareal algorithm using a simplified model instead of the coarse solver. This was done prior to the invention of parareal by Saha et al. \cite{parallelsolar}, also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of Sun and other planets. \cite{parallelsolar} \cite{farfuture}

\subsection{Choice of the coarse and fine integrators}
\begin{itemize}
    \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties with a small cost and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its precision. \cite{integration}
    \item Two symplectic methods could be used, however, due to the nature of the parareal algorithm, the resulting parareal integration would not necessarily be symplectic. A modified parareal algorithm that preserves the symplectic property, shown in \cite{symplecticparareal}, also exists, however it is beyond the scope of this paper.
    \item An adaptive step size fine integrator with error control would be an option, such as the Dormand-Prince RK45 method. However, since the method can have very different execution times (depending on the step size required to reach desired accuracy), which could potentially be an issue when paralellizing the code, RK4 was used for simplicity.
\end{itemize}

@@

\begin{align*}
x(t+\Delta t) &= x(t) + v(t)\Delta t + \frac{1}{2}a(t)\Delta t^2 \\
v(t+\Delta t) &= v(t) + \frac{a(t) + a(t+\Delta t)}{2}\Delta t
\end{align*}
Here $x(t)$, $v(t)$ and $a(t)$ represent approximations of the corresponding values (position, velocity, acceleration) at timestep $t$, $\Delta t$ is the time step size. Note that only one evaluation of $a$ per step suffices, because we can keep $a(t)$ from the last step. \cite{forverlet}

\subsubsection{The classic fourth-order Runge-Kutta method (RK4)}
\begin{align*}
k_1 &= f(t, u(t)) \\
k_2 &= f(t + \frac{\Delta t}{2}, u(t) + \Delta t\frac{k_1}{2}) \\
k_3 &= f(t + \frac{\Delta t}{2}, u(t) + \Delta t\frac{k_2}{2}) \\
k_4 &= f(t + \Delta t, u(t) + \Delta t k_3) \\
u(t+\Delta t) &= u(t) + \frac{\Delta t}{6}\left(k_1 + 2k_2 + 2k_3 + k_4\right)
\end{align*}
where $u(t)$ represents the approximation of the state at time $t$. $\Delta t$ denotes the time step size. The method requires four acceleration computations per step. \cite{forrk4}

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_2.eps}}
\caption{A more complex example to illustrate the behavior of the parareal algorithm: Path of Mercury over the time of one Earth year in a parareal integrated simulation converging over multiple iterations of the parareal algorithm. Velocity verlet and RK4 are used with the same step size of $0.5\text{ day}$.}
\label{mercury}
\end{figure}

@@

\subsection{Experiments: The environment}
All experiments were ran on a cluster in order to get a more reproducible and comparable environment.

@@

Multiple experiments with the same initial conditions and same step size were conducted. A very small step size was used, so all of the runs would converge within the same number of iterations (one iteration\footnote{This choice is explained in the next experiment (\textit{Experiment II}), where the observations suggest that convergence within one iteration possibly has optimal run time.}). For comparison, also a serial version (not using the parareal algorithm) was ran with the same initial conditions and step size.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_4.eps}}
\caption{Conservation of energy in the simulation. Same simulation as in Figure \ref{sim}. With a small enough step size, even a non-symplectic integrator like RK4 can preserve energy for some number of timesteps.}
\label{energy}
\end{figure}

Let's derive the expected speed-ups and run times of our program and compare it with the measured run times. To simplify the calculation, we will only estimate it using the number of function evaluations. For Velocity Verlet, we have $T_\mathcal{C} = 1N_\mathcal{C}$, where $N_\mathcal{C}$ is the number of steps of the coarse integrator. RK4 takes four function evaluations, so we have $T_\mathcal{F} = 4N_\mathcal{F}$. If the fine integrator does 10 times more steps than the coarse one, we have $\frac{N_\mathcal{C}}{N_\mathcal{F}} = \frac{1}{10}$. Thus, we have
$$
S_\text{parareal} = \left(\frac{k}{N} + (k+1)\frac{1N_\mathcal{C}}{4N_\mathcal{F}}\right)^{-1} = \left(\frac{k}{N} + \frac{k+1}{40}\right)^{-1}
$$
When we choose the number of processors (and the number of segments), $N = 16$ and we converge within one iteration, we have
$$
S_\text{parareal} = \left(\frac{1}{16} + \frac{1}{20}\right)^{-1} \approx 8.89
$$

We can compare this with the real measured speed-up, see Table \ref{runtimes}. We observe that in reality the program runs around 5 times faster with 16 threads. We can also compare the expected speed-up $S_\text{parareal}$ for other numbers of CPUs, see Figure \ref{speeds}.

@@

While we loose some accuracy with the parareal method, this is a parameter that can be controlled, for example using the converge parameter ($\varepsilon$) and the coarse solver accuracy.

@@

The second experiment shows how do run times and the numbers of iterations that the program takes to converge relate to the number of steps of the coarse solver. This observation can help with deciding what is the optimal number of steps to use for the coarse solver. We are interested primarily in the run time, however the number of iterations turns out to be a very important factor in the run time.

A series of simulations with the same configuration but with different numbers of steps of the coarse solver ($N_\mathcal{C}$) were ran and the results are shown in Figure \ref{timetest}. The parameter for the stopping criterion was fixed at $\varepsilon = 10^{-3}$ and we observe varying numbers of iterations required to reach this criterion.

@@

From this, we can conjecture, that (in the case of our simulation) the optimal choice of the number of steps for the coarse solver (in the sense of the smallest run time with a fixed $\varepsilon$) is the smallest possible number such that parareal converges within one iteration (or in terms of step sizes, optimal choice of the step size is the biggest possible such that the algorithm converges in one iteration).

@@

\begin{table*}[htbp]
\caption{Comparison of running times taken by different configurations of the simulator for a planetary simulation of $10^4$ days.
$N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively (within one segment). The fine solver step numbers have been chosen specifically to preserve the step size of 0.003125 day. This step size was selected, so all the parareal runs converge within one iteration (with $\varepsilon = 10^{-3}\text{ AU}$). This number of iterations was chosen for easier comparison between the runs and to achieve greater speed-up (see \textit{Experiment II}).
The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
\begin{center}
\begin{tabular}{crrrrrrrrrcc}
\toprule
Program & $n$ CPUs $^{\mathrm{a}}$ & $N_\mathcal{C}$ & $N_\mathcal{F}$ & $\Delta t_\mathcal{F}$ $^{\mathrm{b}}$ & Abs. err. $^{\mathrm{c}}$ & Rel. err. $^{\mathrm{d}}$ & CPU time & Real time \\
\midrule
%\multicolumn{11}{c}{Solar system $10^4\text{days}$} \\
%\midrule[.025em]
parareal & 64 & $0.5\times10^4$ &  $0.5\times10^5$ & 0.003125 day & 0.0110 AU & 3.60\% & 00:11:53 & 00:00:35 \\
parareal & 32 & $1\times10^4$ &  $1\times10^5$ & 0.003125 day & 0.0110 AU & 3.60\% & 00:12:53 & 00:00:52 \\
parareal & 16 & $2\times10^4$ &  $2\times10^5$ & 0.003125 day & 0.0109 AU & 3.60\% & 00:10:25 & 00:01:01 \\
parareal & 4 & $8\times10^4$ &  $8\times10^5$ & 0.003125 day & 0.0109 AU & 3.57\% & 00:10:27 & 00:02:54 \\
serial (base) & 1 & N/A & $32\times10^5$ & 0.003125 day & 0 AU & 0\% & 00:05:01 & 00:05:01 \\
serial (approx.) $^{\mathrm{e}}$ & 1 & N/A & $0.9\times32\times10^5$ & 0.003472 day & 6.7580 AU & 256.25\% & 00:04:38 & 00:04:38 \\
\bottomrule
\multicolumn{8}{l}{$^{\mathrm{a}}$ The number of segments ($N$) is the same as the number of CPUs.} \\
\multicolumn{8}{l}{$^{\mathrm{b}}$ The resulting step size (fine solver step size). For parareal, it can be calculated as $\frac{T-t_0}{NN_\mathcal{F}}$} \\
\multicolumn{8}{l}{$^{\mathrm{c}}$ Maximum difference from the base serial computation across the trajectory} \\
\multicolumn{8}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the base serial computation} \\
\multicolumn{8}{l}{$^{\mathrm{e}}$ A serial approximation of the base serial simulation using a bigger step size.}
\end{tabular}
\label{runtimes}
\end{center}
\end{table*}

@@

The implementation is consistently slower than the estimated run times, however, there are many factors that weren't considered in the speed-up estimation, for example we only took the function evaluations in the solvers into account. There are also other factors such as the synchronization costs of the OpenMP threads.

In conclusion, we were able to achieve noticable speed-ups consistent with the theoretical foundation by using the parareal algorithm.

@@

And finally in Experiment II, we attempted to find the optimal setting (with the smallest run time and a fixed $\varepsilon$) of the course solver step size for our simulation. The results are presented in Figure \ref{timetest}. The result is a prediction of the optimal coarse solver step size for our simulation, which is the biggest possible such that the algorithm converges in one iteration.

@@

In some situations it is not that clear as in our case whether is it better to use spatial or temporal parallelization or both. This is beyond the scope of this article, however, there has been research in this area.

For example, Croce et al. presented a study using a simulation of unsteady Navier-Stokes equations for incompressible flow using parareal that is parallel both in time and space. They measured the speed-up with different numbers of cores being asigned to the spatial and temporal parallelization and shown that they can be combined to gain higher speed-up. \cite{spacetime}

@@

We evaluated parareal only in the example of our solar system. Those settings, however, have their own specifics, in particular, planets in our solar system follow orbits with very different periods, which makes them harder to simulate in the sense that if we simulate a single orbit of Neptune, we have to simulate many orbits of Mercury (in a simulation of all the planets), which requires very small step sizes to behave correctly.

We could further improve our solar system simulations by further specializing them as was done in the work by Saha et al. \cite{parallelsolar} (a different coarse model simulating only the interactions of Sun and other planets was used instead of a course integrator). However, this would be too specific for solar system dynamics and wound not generalize well, so this is beyond our scope.

@@

I would like to thank my advisor, Samuel J. Newcome, M. Sc. and all other participants of the Seminar in Methods for Molecular Dynamics, for suggesting directions, providing valueable feedback and organising this amazing seminar.

@@

\IEEEtriggeratref{9}

@@

\bibitem{parareal}Lions, J., Maday, Y. \& Turinici, G. Résolution d'EDP par un schéma en temps «pararéel». {\em Comptes Rendus De L'Académie Des Sciences - Series I - Mathematics}. \textbf{332}, 661-668 (2001), [Online]. Available: \url{https://www.sciencedirect.com/science/article/pii/S0764444200017936}

@@

\bibitem{farfuture}Gander, M. Is it possible to predict the far future before the near future is known accurately?. {\em Snapshots Of Modern Mathematics From Oberwolfach; 2019}. pp. 21 (2019), [Online]. Available: \url{http://publications.mfo.de/handle/mfo/3693}

>>>>

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{comment}
\usepackage[thinc]{esdiff}
\usepackage{hyperref}

@@

\author{\IEEEauthorblockN{Lukáš Veškrna}
\IEEEauthorblockA{\textit{Technical University Munich}\\ \textit{\& Charles University (Prague)} \\
Munich, Germany \\
lukas.veskrna@gmail.com}
}

@@

\begin{abstract}
The purpose of this paper is to examine the behavior of parallel-in-time integration in particle simulations, in particular the parareal algorithm, by developing a simple simulator for solar system dynamics that utilizes parallel-in-time integration. We present the results of experiments measuring the run times of the simulation, which we then compare with expectations derived from theory.
\end{abstract}

@@

To study the behavior of mostly stable planetary orbits, long simulations are required. Thus, parallel-in-time algorithms provide a promising approach. \cite{parallelsolar}

@@

We typically solve this problem using a numerical method, that does a step in time of size $\Delta t$ and estimates $u(t_0+\Delta t)$. We then use this result to advance the computation further in time by taking more such steps.\footnote{One such method is the explicit Euler method, which simply approximates $u$ at time $t+\Delta t$ by taking a step in the direction given by the value of $f$ at the current state $u$ at time $t$, so $u(t+\Delta t) \approx u(t) + \Delta t f(t, u(t))$. \cite{forrk4}}

@@

Suppose we integrate from time $t_0$ to $T$, we split this interval into $N$ segments of size $\Delta t = \frac{T-t_0}{N}$. We denote $u_0, u_1\dots u_N$ the states of the system in times $t_0, t_1\dots t_N$ with $t_n = t_0 + n\Delta t$ (so $t_0 = t_0$ and $t_N = T$).\footnote{The solvers can also use their own smaller subdivisions (e.g. different step sizes) but the parareal algorithm does not interact with those subdivisions.}

@@

This is repeated until satisfactory convergence. It has been proven for example by Gander et al. \cite{parareal2} that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta t, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.

In the work by Gander et al. \cite{parareal2}, the method has been formulated using different generalized frameworks. They also extend the time interval partitioning to arbitrary partitions, not only equally sized ones as in the original paper. \cite{parareal}

@@

A derivation of the theoretical speedup is presented in Pentland et al. \cite{speedup} (Section 2.3), which we use here. It calculates the speed-up when running even the coarse solver partially in parallel, we simplify it by running $\mathcal{C}$ serially (our implementation doesn't include this optimization).

@@

It is also possible to use a similar algorithm to the parareal algorithm using a simplified model instead of the coarse solver. This was done prior to the invention of parareal by Saha et al. \cite{parallelsolar}, also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of the Sun and other planets. \cite{parallelsolar} \cite{farfuture}

\subsection{Choice of the coarse and fine integrators}
\begin{itemize}
    \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties with a small cost and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its precision. \cite{integration}
    \item Two symplectic methods could be used\footnote{A symplectic integrator preserves physically meaningful behavior of a Hamiltonian system in the long term. \cite{integration}}, however, due to the nature of the parareal algorithm, the resulting parareal integration would not necessarily be symplectic. A modified parareal algorithm that preserves the symplectic property, shown in \cite{symplecticparareal}, also exists, however it is beyond the scope of this paper.
    \item An adaptive step size fine integrator with error control would be an option, such as the Dormand-Prince RK45 method. However, since the method can have very different execution times (depending on the step size required to reach the desired accuracy), which could potentially be an issue when parallelizing the code, RK4 was used for simplicity.
\end{itemize}

@@

\begin{align*}
x(t+\Delta t) &\approx x(t) + v(t)\Delta t + \frac{1}{2}a(t)\Delta t^2 \\
v(t+\Delta t) &\approx v(t) + \frac{a(t) + a(t+\Delta t)}{2}\Delta t
\end{align*}
Here $x(t)$, $v(t)$ and $a(t)$ are approximations of the corresponding values (position, velocity, acceleration) by the method at time $t$, $\Delta t$ is the time step size. Note that only one evaluation of $a$ per step suffices, because we can keep $a(t)$ from the last step. \cite{forverlet}

\subsubsection{The classic fourth-order Runge-Kutta method (RK4)}
\begin{align*}
k_1 &= f(t, u(t)) \\
k_2 &= f(t + \frac{\Delta t}{2}, u(t) + \Delta t\frac{k_1}{2}) \\
k_3 &= f(t + \frac{\Delta t}{2}, u(t) + \Delta t\frac{k_2}{2}) \\
k_4 &= f(t + \Delta t, u(t) + \Delta t k_3) \\
u(t+\Delta t) &\approx u(t) + \frac{\Delta t}{6}\left(k_1 + 2k_2 + 2k_3 + k_4\right)
\end{align*}
where $u(t)$ is the approximation of the state by the method at time $t$. $\Delta t$ denotes the time step size. The method requires four acceleration computations per step. \cite{forrk4}

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_2.eps}}
\caption{A more complex example to illustrate the behavior of the parareal algorithm: Path of Mercury over the time of one Earth year in a parareal integrated simulation converging over multiple iterations of the parareal algorithm. Velocity Verlet and RK4 are used with the same step size of $0.5\text{ day}$.}
\label{mercury}
\end{figure}

@@

\subsection{Experiments: The environment}
All experiments were run on a cluster in order to get a more reproducible and comparable environment.

@@

Multiple experiments with the same initial conditions and the same step size were conducted. A very small step size was used, so all of the runs would converge within the same number of iterations (one iteration\footnote{This choice is explained in the next experiment (\textit{Experiment II}), where the observations suggest that convergence within one iteration possibly has optimal run time.}). For comparison, also a serial version (not using the parareal algorithm) was run with the same initial conditions and step size.

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_4.eps}}
\caption{Conservation of energy in the simulation. Same simulation as in Figure \ref{sim}. With a small enough step size, even a non-symplectic integrator like RK4 can preserve energy for some number of time steps.}
\label{energy}
\end{figure}

Let's derive the expected speed-ups and run times of our program and compare them with the measured run times. To simplify the calculation, we will only estimate it using the number of function evaluations. For Velocity Verlet, we have $T_\mathcal{C} = 1N_\mathcal{C}$, where $N_\mathcal{C}$ is the number of steps of the coarse integrator. RK4 takes four function evaluations, so we have $T_\mathcal{F} = 4N_\mathcal{F}$. If the fine integrator does 10 times more steps than the coarse one, we have $\frac{N_\mathcal{C}}{N_\mathcal{F}} = \frac{1}{10}$. Thus, we have
$$
S_\text{parareal} = \left(\frac{k}{N} + (k+1)\frac{1N_\mathcal{C}}{4N_\mathcal{F}}\right)^{-1} = \left(\frac{k}{N} + \frac{k+1}{40}\right)^{-1}
$$
When we choose the number of processors (and the number of segments), $N = 16$ and we converge within one iteration, we have
$$
S_\text{parareal} = \left(\frac{1}{16} + \frac{1}{20}\right)^{-1} \approx 8.89
$$

We can compare this with the real measured speed-up, see Table \ref{runtimes}. We observe that in reality, the program runs around 5 times faster with 16 threads. We can also compare the expected speed-up $S_\text{parareal}$ for other numbers of CPUs, see Figure \ref{speeds}.

@@

While we lose some accuracy with the parareal method, this is a parameter that can be controlled, for example using the converge parameter ($\varepsilon$) and the coarse solver accuracy.

@@

The second experiment shows how run times and the numbers of iterations that the program takes to converge relate to the number of steps of the coarse solver. This observation can help with deciding what is the optimal number of steps to use for the coarse solver. We are interested primarily in the run time, however the number of iterations turns out to be a very important factor in the run time.

A series of simulations with the same configuration but with different numbers of steps of the coarse solver ($N_\mathcal{C}$) were run and the results are shown in Figure \ref{timetest}. The parameter for the stopping criterion was fixed at $\varepsilon = 10^{-3}$ and we observe varying numbers of iterations required to reach this criterion.

@@

From this, we can conjecture, that (in the case of our simulation) the optimal choice of the number of steps for the coarse solver (in the sense of the smallest run time with a fixed $\varepsilon$) is the smallest possible number such that parareal converges within one iteration (or in terms of step sizes, the optimal choice of the step size is the biggest possible such that the algorithm converges in one iteration).

@@

\begin{table*}[htbp]
\caption{Comparison of running times taken by different configurations of the simulator for a planetary simulation of $10^4$ days.
$N_\mathcal{C}$ and $N_\mathcal{F}$ refer to the number of steps of the coarse and the fine integrators respectively (within one segment). The fine solver step numbers have been chosen specifically to preserve the step size of 0.003125 days. This step size was selected, so all the parareal runs converge within one iteration (with $\varepsilon = 10^{-3}\text{ AU}$). This number of iterations was chosen for easier comparison between the runs and to achieve greater speed-up (see \textit{Experiment II}).
The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equipped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
\begin{center}
\begin{tabular}{crrrrrrrrrcc}
\toprule
Program & $n$ CPUs $^{\mathrm{a}}$ & $N_\mathcal{C}$ & $N_\mathcal{F}$ & $\Delta t_\mathcal{F}$ $^{\mathrm{b}}$ & Abs. err. $^{\mathrm{c}}$ & Rel. err. $^{\mathrm{d}}$ & CPU time & Real time \\
\midrule
%\multicolumn{11}{c}{Solar system $10^4\text{days}$} \\
%\midrule[.025em]
parareal & 64 & $0.5\times10^4$ &  $0.5\times10^5$ & 0.003125 day & 0.0110 AU & 3.60\% & 00:11:53 & 00:00:35 \\
parareal & 32 & $1\times10^4$ &  $1\times10^5$ & 0.003125 day & 0.0110 AU & 3.60\% & 00:12:53 & 00:00:52 \\
parareal & 16 & $2\times10^4$ &  $2\times10^5$ & 0.003125 day & 0.0109 AU & 3.60\% & 00:10:25 & 00:01:01 \\
parareal & 4 & $8\times10^4$ &  $8\times10^5$ & 0.003125 day & 0.0109 AU & 3.57\% & 00:10:27 & 00:02:54 \\
serial (base) & 1 & N/A & $32\times10^5$ & 0.003125 day & 0 AU & 0\% & 00:05:01 & 00:05:01 \\
serial (approx.) $^{\mathrm{e}}$ & 1 & N/A & $0.9\times32\times10^5$ & 0.003472 day & 6.7580 AU & 256.25\% & 00:04:38 & 00:04:38 \\
\bottomrule
\multicolumn{8}{l}{$^{\mathrm{a}}$ The number of segments ($N$) is the same as the number of CPUs.} \\
\multicolumn{8}{l}{$^{\mathrm{b}}$ The resulting step size (fine solver step size). For parareal, it can be calculated as $\frac{T-t_0}{NN_\mathcal{F}}$} \\
\multicolumn{8}{l}{$^{\mathrm{c}}$ Maximum difference from the base serial computation across the trajectory} \\
\multicolumn{8}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the base serial computation} \\
\multicolumn{8}{l}{$^{\mathrm{e}}$ A serial approximation of the base serial simulation using a bigger step size.}
\end{tabular}
\label{runtimes}
\end{center}
\end{table*}

@@

The implementation is consistently slower than the estimated run times; however, many factors were not considered in the speed-up estimation, for example, in the solvers we only took into account the function evaluations. There are also other factors such as the synchronization costs of the OpenMP threads.

In conclusion, we were able to achieve noticeable speed-ups consistent with the theoretical foundation by using the parareal algorithm.

@@

Finally, in Experiment II, we attempted to find the optimal setting (with the smallest run time and a fixed $\varepsilon$) of the course solver step size for our simulation. The results are presented in Figure \ref{timetest}. The result is a prediction of the optimal coarse solver step size for our simulation, which is the biggest possible such that the algorithm converges in one iteration.

@@

In some situations, it is not that clear as in our case whether is it better to use spatial or temporal parallelization or both. This is beyond the scope of this article, however, there has been research in this area.

For example, Croce et al. presented a study using a simulation of unsteady Navier-Stokes equations for incompressible flow using parareal that is parallel both in time and space. They measured the speed-up with different numbers of cores assigned to the spatial and temporal parallelization and showed that they can be combined to gain higher speed-up. \cite{spacetime}

@@

We evaluated parareal only in the example of our solar system. Those settings, however, have their own specifics, for example, planets in our solar system follow orbits with very different periods, which makes them harder to simulate in the sense that if we simulate a single orbit of Neptune, we have to simulate many orbits of Mercury (in a simulation of all the planets), which requires very small step sizes to behave correctly.

We could further improve our solar system simulations by further specializing them as was done in the work by Saha et al. \cite{parallelsolar} (a different coarse model simulating only the interactions of the Sun and other planets was used instead of a course integrator). However, this would be too specific for solar system dynamics and wound not generalize well, so this is beyond our scope.

@@

I would like to thank my advisor, Samuel J. Newcome, M. Sc. and all other participants of the Seminar in Methods for Molecular Dynamics, for suggesting directions, providing valuable feedback and organizing this amazing seminar.

@@

\IEEEtriggeratref{7}

@@

\bibitem{parareal}Lions, J., Maday, Y. \& Turinici, G. Résolution d'EDP par un schéma en temps «pararéel». {\em Comptes Rendus De L'Académie Des Sciences - Series I - Mathematics}. \textbf{332}, 661-668 (2001), [Online]. Available: \url{https://doi.org/10.1016/S0764-4442(00)01793-6}

@@

\bibitem{farfuture}Gander, M. Is it possible to predict the far future before the near future is known accurately?. {\em Snapshots Of Modern Mathematics From Oberwolfach; 2019}. pp. 21 (2019), [Online]. Available: \url{https://doi.org/10.14760/SNAP-2019-021-EN}
